---
title: "GWASTargetChase Package Documentation"
author: "Arby Abood"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GWASTargetChase Package Documentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Purpose

Complex diseases are disorders that results from multiple genetic variants and genes coupled with influences from the environment. One way to study complex disease is Genome-wide association studies (GWAS). GWAS is an approach to study cohorts (populations) and associate loci (genomic regions containing multiple variants and genes) to complex diseases. However, a main challenge to interpreting GWAS results is 90% of these loci are found within intronic or intergenic regions suggesting they are involved in gene regulation rather than affecting the gene sequence. Therefore, GWAS follow-up (i.e.using multi-omics data and approaches to pin point causal genes) has been used as a tool in human genetics to identify causal genes in complex diseases and therefore hone in on potential therapeutic and intervention targets.

Research funding has been poured into generating resources of GWAS follow-up data in humans given the success this approach has shown in identifying gene targets. However, other species are lagging behind due to the lack of funding. Therefore, the idea to use humans as a model system for feline complex disease could potentially point to genes of interest to feline health. Thus, the purpose of this package is to prepare, and utilize data from human studies to follow up on feline and/or K9 GWAS summary statistics.

# Data Curation

The human GWAS follow-up data is obtained from the following sources:

- OpenTargets <https://www.opentargets.org/>
- IMPC <https://www.mousephenotype.org/>

Cross-species gene translation is performed using Zoonomia orthology data included with the package.

# Installation and Loading

```{r}
# Install dependencies first
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# Install Bioconductor packages
BiocManager::install(c("GenomicRanges", "rtracklayer", "S4Vectors"))

# Install CRAN packages
install.packages(c("data.table", "dplyr", "arrow", "jsonlite",
                   "ggplot2", "ggpubr", "patchwork", "remotes"))

# Install GWASTargetChase from GitHub
remotes::install_github("aa9gj/GWASTargetChase")

# Load the package
library(GWASTargetChase)
```

# Recommended GTF files

Download the following GTF files. Please note this package depends on Ensemble data structure for gene names and chromosome IDs.

```{sh}
wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_40/gencode.v40.annotation.gtf.gz
gunzip gencode.v40.annotation.gtf.gz

wget http://ftp.ensembl.org/pub/release-106/gtf/felis_catus/Felis_catus.Felis_catus_9.0.106.gtf.gz
gunzip Felis_catus.Felis_catus_9.0.106.gtf.gz
```

# Functions

There are two main functions within this package: `TargetChase` and `TargetChaseManual`. Additional helper functions include `IMPCprep`, `l2gPrep`, `geneticAssocPrep`, and `downloadData`.

Please ensure that your sumStats file has the following column names exactly as such: (The rest can stay as is)

- `p_wald`: This would be the corrected p-values
- `ps`: SNP position
- `chr`: chromosome
- `gene`: closest gene name per locus (required for `TargetChase`)

## TargetChase

This is the main function of this package. It takes GWAS summary stats and a GTF file. It queries the OpenTargets API for gene-disease associations and the IMPC API for mouse knockout phenotypes. For non-human species (cat, dog), it uses Zoonomia orthology to translate genes to human/mouse orthologs.

```{r}
TargetChase(sumStats = "/path/to/GWAS/Summary/stats",
            gtf = "/path/to/feline/or/k9/gtf/file",
            species = "cat",
            pval = 0.00005,
            ResultsPath = "/path/to/results/directory/")
```

## TargetChaseManual

Use this function when you have pre-downloaded OpenTargets and IMPC data files. This is useful when you have updated data releases or want to avoid repeated API calls.

Once the data has been prepared by the helper functions below (or via `downloadData`), you can use:

```{r}
TargetChaseManual(sumStats = "/path/to/sumstats",
                  gtf = "/path/to/Felis_catus.Felis_catus_9.0.106.gtf",
                  species = "cat",
                  pval = 0.00005,
                  ResultsPath = "/path/to/results/",
                  impc = "/path/to/impc/results",
                  assocOT = "/path/to/disease_target_genetic_associations",
                  l2gOT = "/path/to/l2g_annotated_full")
```

## downloadData

Downloads and prepares all required data files (IMPC, OpenTargets associations, L2G scores, GENCODE GTF) in one step:

```{r}
paths <- downloadData(destdir = "my_data")

# Then use with TargetChaseManual:
TargetChaseManual(
  sumStats = "my_gwas.txt",
  gtf = "my_species.gtf",
  impc = paths$impc,
  assocOT = paths$assoc,
  l2gOT = paths$l2g
)
```

## IMPCprep

You would need to download the relevant IMPC data from the command line:

```{sh}
wget ftp://ftp.ebi.ac.uk/pub/databases/impc//all-data-releases/release-16.0/results/phenotypeHitsPerGene.csv.gz
gunzip phenotypeHitsPerGene.csv.gz
```

In R/Rstudio, use the following command to prepare the IMPC data for further analysis:

```{r}
IMPCprep(impcData = "/path/to/phenotypeHitsPerGene.csv",
         ResultsPath = "/path/to/results/")
```

## l2gPrep

If you don't want to use the lazyloaded data (in case of an updated release), you'll have to prepare the data yourself. This function prepares locus2gene data from OpenTargets (OT). You would need to download the l2g and study index data from OT using the following commands in the command line.

```{sh}
wget ftp://ftp.ebi.ac.uk/pub/databases/opentargets/genetics/latest/l2g/*
wget ftp://ftp.ebi.ac.uk/pub/databases/opentargets/genetics/latest/lut/study-index/*
cat *.json > study_index.json
```

In R/Rstudio, use the following command to prepare the locus2gene data for further analysis:

```{r}
l2gPrep(l2gOT = "/path/to/l2g/dir/l2g/",
        studyOT = "/path/to/study_index.json",
        humanGTF = "/path/to/humangtf/e.g.gencode.v40.annotation.gtf.gz",
        ResultsPath = "/path/to/results/dir/")
```

## geneticAssocPrep

If you don't want to use the lazyloaded data (in case of an updated release), you'll have to prepare the data yourself. This function prepares the target data from OpenTargets (OT). This information is not as reliable as their locus2gene model. You would need to download the target-disease associations (direct) and disease annotation data from OT using the following commands in the command line.

```{sh}
wget --recursive --no-parent --no-host-directories --cut-dirs 8 ftp://ftp.ebi.ac.uk/pub/databases/opentargets/platform/22.06/output/etl/json/associationByDatasourceDirect
cat *.json > associationByDatasourceDirect.json

wget --recursive --no-parent --no-host-directories --cut-dirs 8 ftp://ftp.ebi.ac.uk/pub/databases/opentargets/platform/22.06/output/etl/json/diseases
cat *.json > diseases.json
```

In R/Rstudio, use the following command to prepare the genetic association data for further analysis:

```{r}
geneticAssocPrep(assocOT = "/path/to/dir/associationByDatasourceDirect.json",
                 diseaseOT = "/path/to/dir/diseases.json",
                 humanGTF = "/path/to/e.g.gencode.v40.annotation.gtf.gz",
                 ResultsPath = "/path/to/res/")
```

# Output

If you use `TargetChase` function, here is the output you would get:

- **g2d_results**: A score of how "targetable" the target is from OpenTargets
- **impc_results**: IMPC mouse knockout phenotype data
- **plots**: LocusZoom plots for all significant loci (p_wald > 5)

If you use `TargetChaseManual`, you will additionally get:

- **l2g_results**: Full locus-to-gene results from OpenTargets
